Gametic Selection, Sex Ratio Bias, and Transitions Between Sex Determination Systems

### 2017-01-19 LaTeX/Turnover.tex
Adding extra results parts

@@

%genetic system
We consider the transition between an ancestral and novel sex determination systems using a three locus model.
Locus \textbf{X} is the ancestral sex-determining region, with alleles $X$ and $Y$ (or $Z$ and $W$).
Locus \textbf{A} is a locus under selection, with alleles $A$ and $a$.
Locus \textbf{M} is a novel sex-determining region, at which the null allele ($M$) is initially fixed in the population such that sex of zygotes is determined by the genotype at the ancestral sex-determining region, \textbf{X} ($XX$ become females and $XY$ become males, or $ZW$ become females and $ZZ$ become males). 
To evaluate the evolution of new sex-determination systems, we consider the invasion, fixation, maintenance, and/or loss of novel sex-determining alleles ($m$) at the \textbf{M} locus. 
%We only do invasion analytically though (is this sentence still ok?)
We assume that the \textbf{M} locus is dominant over the \textbf{X} locus such that zygotes with at least one $m$ allele at locus \textbf{M}, a zygote develops as a female with probability $k$ and as a male with probability $1-k$, regardless of the \textbf{X} locus genotype.
With $k=0$, the $m$ allele is a masculinizer (i.e., a neo-Y) and with $k=1$ the $m$ allele is a feminizer (i.e., a neo-W).
With intermediate $k$, the $m$ allele confers environmental sex determination such that zygotes develop as females in a proportion ($k$) of the environments they experience. 
Finally, we also analyze a model of maternally-controlled environmental sex-determination (ESD), where mothers with at least one $m$ allele produce daughters with probability $k$.

%life-cycle
In each generation, we census the genotype frequencies in male and female gametes/gametophytes (hereafter gametes) before haploid competition (recursion equations in Sup.\ Mat.). 
First, competition occurs among male gametes (sperm/pollen competition) and among female gametes (egg/ovule competition). 
Selection during haploid competition depends on the \textbf{A} locus genotype, fitnesses are $w_{A}^m$ and $w_{a}^m$ for male gametes and $w_{A}^f$ and $w_{a}^f$ for female gametes, see table \ref{tab:fitnesstable}. 
Random mating then occurs between male and female gametes.
The resulting zygotes develop as males or females, depending on their genotypes at the \textbf{X} and \textbf{M} loci (and the \textbf{M} genotype of their mother in the case of maternal control) as described above.
Diploid males and females then experience selection, male fitness is given by $w_{h}^{m}$ and female fitness by $w_{h}^{f}$, where $h$ is the genotype at the \textbf{A} locus ($h \in AA, Aa, aa$).  
%Diploids compete with others of the same sex, where selection depends on the sex of the individual and its genotype at the \textbf{A} locus.
The next generation of gametes are then produced by meiosis, during which recombination and sex-specific meiotic drive can occur. 
Recombination occurs between loci \textbf{X} and \textbf{A} with probability $r$, between loci \textbf{A} and \textbf{M} with probability $R$, and between loci \textbf{X} and \textbf{M} with probability $\chi$.
Therefore, any order of the loci can be modelled with appropriate choices of $r$, $R$, and $\chi$ (see Table \ref{tab:chisubstitutions}). 
Males/females that are heterozygous at the \textbf{A} locus experience meiotic drive; $Aa$ heterozgotes of sex $d$ produce gametes bearing allele $A$ with probability $\alpha^d$. 
Thus, the \textbf{A} locus experience sex-specific haploid competition, diploid selection and/or meiotic drive.

\begin{table}[ht]
\centering
\smallskip
\caption{Fitness of different genotypes in sex $d$}
\begin{tabular}{l c }
\hline\hline
  Genotype & Fitness \\ [0.5ex] \hline
  Haploid Competition & \\
  A & $w_{A}^d = 1+t^d$ \\
  a & $w_{a}^d = 1$ \\
  Diploid Selection & \\
  AA & $w_{AA}^d = 1+ s^d$ \\
  Aa & $w_{Aa}^d = 1+h^k s^d$ \\
  aa & $w_{aa}^d = 1$ \\
  \hline \hline
  \label{tab:fitnesstable}
 \end{tabular}
\end{table}

@@

Various forms of selection can maintain a polymorphism at the \textbf{A} locus, including sexually antagonistic selection, overdominance and conflicts between diploid selection and selection upon haploid genotypes \citep[ploidally antagonistic selection,][]{Immler:2012tl} or a combination of these selective regimes. 
Here, we assume that selection and meiotic drive are weak relative to recombination ($s^k$ and $t^k$ of order $\epsilon$ and $\alpha^k=1/2+\alpha_{\Delta}^k$, where $\alpha_{\Delta}^k$ is of order $\epsilon$). 
The maintenance of a polymorphism at the \textbf{A} locus then requires that

@@

\begin{equation}
\begin{split}
\hat{p}^m_X-\hat{p}^f_X&=V_{A}\big{(}C^m - C^f \big{)}
+O(\epsilon^2)\\
%
\hat{p}^m_Y-\hat{p}^f_X&=V_{A}\big{(}C^m - C^f+(1-2r)(t^m-t^f)\big{)}/2r
+O(\epsilon^2)\\
%
\hat{p}^m_Y-\hat{p}^m_X&=V_{A}\big{(}C^m - C^f+(t^m-t^f)\big{)}(1-2r)/2r
+O(\epsilon^2)
\end{split}
\end{equation}

\noindent
where $V_{A}=\bar{p}(1-\bar{p})$ is the variance in the frequency of $A$ and $C^d=(\bar{p}+(1-2\bar{p})h^d) s^d +2\alpha_{\Delta}^d$. 
The frequency of Y among male gametes is proportional to the difference in $A$ allele frequency on X- and Y-bearing male gametes, $q=\alpha_{\Delta}^m(\hat{p}^m_Y-\hat{p}^m_X)+O(\epsilon^3)$.
Without haploid competition or drive ($\alpha_{\Delta}^d=t^d=0$), these results reduce to those of \citet{vanDoorn:2007eu}.

@@

\subsection*{Sex chromosome turnover}
The evolution of a new sex determination system requires that a rare mutant, $m$, at the novel sex-determining locus  increases in frequency when rare. 
The spread of a rare mutant $m$ at the \textbf{M} locus is determined by the leading eigenvalue, $\lambda$, of the system described by the next generation frequency of eggs and sperm carrying the mutation, (\ref{eq:recursions}c), (\ref{eq:recursions}d), (\ref{eq:recursions}g), (\ref{eq:recursions}h), which is an eight equation system. %, evaluated at the above equilibrium.
%Thus, the system describing the change in frequency of the $m$ allele consists of eight recursion equations. 
Dominant neo-Y chromosomes (when $k=0$) or neo-W chromosomes (when $k=1$) are only found in male diploids (neo-Y) or female diploids (neo-W) such that their long term growth rate depends only on the change in frequency of $m$-bearing gametes produced by males (for a neo-Y) or by females (for a neo-W). 
Furthermore, if the $m$ allele is fully dominant over the ancestral sex-determining system, phenotypes are not affected by the genotype at the ancestral sex-determining region (\textbf{X} locus). 
Therefore, the invasion of rare mutant neo-Y or neo-W chromosomes can be described by a simpler, two equation system, the largest eigenvalue of which solves the characteristic polynomial

@@

\noindent
where $b$ is the average of the growth rates of the two haplotypes that carry the $m$ allele ($mA$ and $ma$), $b=(\lambda_{mA}^{m}+\lambda_{ma}^{m})/2$, and $c$ also involves the fitness of $m$ alleles when they recombine onto the other \textbf{A} background in a heterozygote, $c=\lambda_{mA}\lambda_{ma}+\rho_{mA}\rho_{ma}$.

\begin{table}[ht]
\centering
\smallskip
\caption{Parameters determining invasion (equation \ref{eq:charpoly_neoY}) for neo-Y or neo-W chromosomes}
\begin{tabular}{l}
\hline\hline
   \noalign{\vskip 0.5ex}
   neo-Y ($k=0$) \\ [0.5ex] \hline \noalign{\vskip 1ex}
  $\lambda_{mA}=\{p_{Xf} w_{A}^{f} w_{A}^{m} w_{AA}^{m}+(1-p_{Xf}) w_{a}^{f} w_{A}^{m} w_{Aa}^{m}\alpha^m(1-R)\}/ \bar{w}_{H}^{f} \bar{w}_{H}^{m} \bar{w}^m$ \\ [0.5ex] \noalign{\vskip 0.5ex}
  $\lambda_{ma}=\{(1-p_{Xf}) w_{a}^{f} w_{a}^{m} w_{aa}^{m}+p_{Xf} w_{A}^{f} w_{a}^{m} w_{Aa}^{m}(1-\alpha^m)(1-R)\}/ \bar{w}_{H}^{f} \bar{w}_{H}^{m} \bar{w}^m $ \\ [0.5ex] \noalign{\vskip 0.5ex}
  $\rho_{mA}=R \{(1-p_{Xf}) w_{a}^{f} w_{A}^{m} w_{Aa}^{m} (1-\alpha_{m})\}/ \bar{w}_{H}^{f} \bar{w}_{H}^{m} \bar{w}^m $\\ [0.5ex] \noalign{\vskip 0.5ex}
  $\rho_{ma}=R\{p_{Xf} w_{A}^{f} w_{a}^{m} w_{Aa}^{m} \alpha_{m}\}/ \bar{w}_{H}^{f} \bar{w}_{H}^{m} \bar{w}^m$\\ [1ex] \hline 
  \noalign{\vskip 0.5ex}
  neo-W ($k=1$) \\ [0.5ex] \hline \noalign{\vskip 1ex}
  $\lambda_{mA}=\{\bar{p}_{m} w_{A}^{m} w_{A}^{f} w_{AA}^{f}+(1-\bar{p}_{m}) w_{a}^{m} w_{A}^{f} w_{Aa}^{f}\alpha^f(1-R)\}/ \bar{w}_{H}^{f} \bar{w}_{H}^{m} \bar{w}^f$ \\ [0.5ex] \noalign{\vskip 0.5ex}
  $\lambda_{ma}=\{(1-\bar{p}_{m}) w_{a}^{m} w_{a}^{f} w_{aa}^{f}+\bar{p}_{m} w_{A}^{m} w_{a}^{f} w_{Aa}^{f}(1-\alpha^f)(1-R)\}/ \bar{w}_{H}^{f} \bar{w}_{H}^{m} \bar{w}^f $ \\ [0.5ex] \noalign{\vskip 0.5ex}
  $\rho_{mA}=R \{(1-\bar{p}_{m}) w_{a}^{m} w_{A}^{f} w_{Aa}^{f} (1-\alpha_{f})\}/ \bar{w}_{H}^{f} \bar{w}_{H}^{m} \bar{w}^f $\\ [0.5ex] \noalign{\vskip 0.5ex}
  $\rho_{ma}=R\{\bar{p}_{m} w_{A}^{m} w_{a}^{f} w_{Aa}^{f} \alpha_{f}\}/ \bar{w}_{H}^{f} \bar{w}_{H}^{m} \bar{w}^f$\\ [1ex] 
  \hline \hline
  \label{tab:haplotype_growth}
 \end{tabular}
\end{table}

@@

%<<<<<<< HEAD
%\subsubsection*{Neo-$Y$}
%
%The growth rate of a rare, dominant neo-$Y$ region ($k=0$) is
%\begin{equation}
%\lambda_{Y,XY} = 1 + V_A \frac{(r - R)}{r R} \frac{s_f^2}{\left( s_f + s_m \right)^2} t_m^2,
%\end{equation}
%where $V_A = p_A(1-p_A)$ is the variance at the \textbf{A} locus, $s_f$ and $s_m$ are the respective selection coefficients for $A$ in diploid females and males, and $t_m$ is the selection coefficient for $A$ in gametes/gametophytes from males.
%The neo-Y can therefore invade whenever it is in tighter linkage with the selected locus than is the ancestral sex-determining locus, $r>R$, provided locus \textbf{A} is polymorphic ($V_A>0$) and there is selection among both females ($s_f\neq0$) and gametes/gametophytes from males ($t_m\neq0$).
%=======
%>>>>>>> c89cd9e1d25a76e5bffdd1ee7574884f24c08070
%write in general form without weak selection
A rare, dominant neo-$Y$ ($k=0$) or neo-$W$ ($k=1$) is always expected to invade an ancestral $XY$ system when the average growth rate of the mutant haplotypes ($Am$ and $am$) is positive, $(g_A + g_a)/2 > 0$. %($g_i = where $\lambda_i$ are the two eigenvalues when $R=r=0$; the \textbf{X} locus does not affect the dynamics).
When the growth rates of the mutant haplotypes without recombination ($R=0$) are negative, $g_i^* < 0 \; \forall \; i \in \{A,a\}$, where $g_i < g_i^*$, the new sex-determining allele does not invade.

Otherwise, the new sex-determining allele increases in frequency on one \textbf{A} background and declines on the other, and invasion requires
\begin{equation}\label{eq:neoYR}
R \left[ \frac{p^f_X w^m_a (1-\alpha^m)}{g_A^*} + \frac{(1 - p^f_X) w^m_A \alpha^m}{g_a^*} \right] w^m_{Aa} < \nu^m,
\end{equation}
for the neo-$Y$, and 
\begin{equation}\label{eq:neoWR}
R \left[  \frac{(1 - \bar{p}^m) w^m_{a}}{g_A^*} + \frac{\bar{p}^m w^m_{A}}{g_a^*} \right] w^f_{Aa} < \nu^f
\end{equation}
for the neo-$W$.
Here $\bar{p}^m = (1-q) p^m_X+ q p^m_Y$ is the average frequency of $A$ in gametes produced by males, $w^i_{j}$ is the relative viability fitness of sex $i\in\{m,f\}$ depending on their haploid or diploid genotype at the \textbf{A} locus (with $j\in\{A,a,AA,Aa,aa\}$) and
$\nu^i = 
  p^f_X p^m_{\textbf{X}^i} w^m_{a} w^i_{AA} +
  p^f_X (1-p^m_{\textbf{X}^i}) w^m_{a} w^i_{Aa} +
  (1-p^f_X) p^m_{\textbf{X}^i} w^m_{a} w^i_{Aa} +
  (1-p^f_X) (1-p^m_{\textbf{X}^i}) w^m_a w^i_{aa}$,
with $\textbf{X}^m = Y$ and $\textbf{X}^f = X$, is the mean relative fitness of resident individuals of sex $i$.
%and $g_j^*$ is the growth rate of $X^m$ and $Ym$ haplotypes when on an $A$ ($j=A$)  or $a$ background ($j=a$).
Although male meiotic drive does not explicitly appear in Equation \ref{eq:neoWR}, it does affect the average frequency of $A$ in gametes from males, $\bar{p}^m$, and thus can play a role in neo-W invasion.

>>>>

\usepackage[flushleft]{threeparttable}

@@

%genetic system
We consider the transition between an ancestral and novel sex determination systems using a three locus model.
Locus \textbf{X} is the ancestral sex-determining region, with alleles $X$ and $Y$ (or $Z$ and $W$).
Locus \textbf{A} is a locus under selection, with alleles $A$ and $a$.
Locus \textbf{M} is a novel sex-determining region, at which the null allele ($M$) is initially fixed in the population such that sex of zygotes is determined by the genotype at the ancestral sex-determining region, \textbf{X} ($XX$ become females and $XY$ become males, or $ZW$ become females and $ZZ$ become males). 
To evaluate the evolution of new sex-determination systems, we consider the invasion, fixation, maintenance, and/or loss of novel sex-determining alleles ($m$) at the \textbf{M} locus. 
%We only do invasion analytically though (is this sentence still ok?)
We assume that the \textbf{M} locus is dominant over the \textbf{X} locus such that zygotes with at least one $m$ allele develop as females with probability $k$ and as males with probability $1-k$, regardless of the \textbf{X} locus genotype.
With $k=0$, the $m$ allele is a masculinizer (i.e., a neo-Y) and with $k=1$ the $m$ allele is a feminizer (i.e., a neo-W).
With intermediate $k$, the $m$ allele confers environmental sex determination such that zygotes develop as females in a proportion ($k$) of the environments they experience. 
Finally, we also analyze a model of maternally-controlled environmental sex-determination (ESD), where mothers with at least one $m$ allele produce daughters with probability $k$.

%life-cycle
In each generation, we census the genotype frequencies in male and female gametes/gametophytes (hereafter gametes) before haploid competition (see Sup.\ Mat. for recursion equations). 
First, competition occurs among male gametes (sperm/pollen competition) and among female gametes (egg/ovule competition). 
Selection during haploid competition depends on the \textbf{A} locus genotype, fitnesses are $w_{A}^m$ and $w_{a}^m$ for male gametes and $w_{A}^f$ and $w_{a}^f$ for female gametes, see table \ref{tab:fitnesstable}. 
Random mating then occurs between male and female gametes.
The resulting zygotes develop as males or females, depending on their genotypes at the \textbf{X} and \textbf{M} loci (and the \textbf{M} genotype of their mother in the case of maternal control) as described above.
Diploid males and females then experience selection, male fitness is given by $w_{h}^{m}$ and female fitness by $w_{h}^{f}$, where $h$ is the genotype at the \textbf{A} locus ($h \in AA, Aa, aa$).  
%Diploids compete with others of the same sex, where selection depends on the sex of the individual and its genotype at the \textbf{A} locus.
The next generation of gametes are then produced by meiosis, during which recombination and sex-specific meiotic drive can occur. 
Recombination occurs between loci \textbf{X} and \textbf{A} with probability $r$, between loci \textbf{A} and \textbf{M} with probability $R$, and between loci \textbf{X} and \textbf{M} with probability $\chi$.
Therefore, any order of the loci can be modelled with appropriate choices of $r$, $R$, and $\chi$ (see Table \ref{tab:chisubstitutions}). 
Males/females that are heterozygous at the \textbf{A} locus experience meiotic drive; $Aa$ heterozgotes of sex $d$ produce gametes bearing allele $A$ with probability $\alpha^d$. 
Thus, the \textbf{A} locus can experience sex-specific haploid competition, diploid selection and/or meiotic drive.

\begin{table}[ht]
\centering
\smallskip
\caption{Fitness of different genotypes in sex $d$}
\begin{tabular}{l c }
\hline\hline
  Genotype & Fitness during haploid competition \\ [0.5ex] \hline
  A & $w_{A}^d = 1+t^d$ \\
  a & $w_{a}^d = 1$ \\ [0.5ex] \hline
  Genotype & Fitness during diploid selection \\ [0.5ex] \hline
  AA & $w_{AA}^d = 1+ s^d$ \\
  Aa & $w_{Aa}^d = 1+h^k s^d$ \\
  aa & $w_{aa}^d = 1$ \\ [0.5ex] \hline
  Genotype & Tranmission during meiosis in $Aa$ heterozygotes \\ [0.5ex] \hline
  A & $\alpha^d=1/2+\alpha_{\Delta}^{d}$ \\
  a & $(1-\alpha^d)=1/2-\alpha_{\Delta}^{d}$ \\
  \hline \hline
  \label{tab:fitnesstable}
 \end{tabular}
\end{table}

@@

Various forms of selection can maintain a polymorphism at the \textbf{A} locus, including sexually antagonistic selection, overdominance and conflicts between diploid selection and selection upon haploid genotypes \citep[ploidally antagonistic selection,][]{Immler:2012tl} or a combination of these selective regimes. 
Here, we assume that selection and meiotic drive are weak relative to recombination ($s^k$, $t^k$, $\alpha_{\Delta}^k$ of order $\epsilon$). 
The maintenance of a polymorphism at the \textbf{A} locus then requires that

@@

\begin{equation}
\begin{split}
\hat{p}^m_X-\hat{p}^f_X&=V_{A}\big{(}C^m - C^f \big{)}
+O(\epsilon^2)\\
%
\hat{p}^m_Y-\hat{p}^f_X&=V_{A}\big{(}C^m - C^f+(1-2r)(t^m-t^f)\big{)}/2r
+O(\epsilon^2)\\
%
\hat{p}^m_Y-\hat{p}^m_X&=V_{A}\big{(}C^m - C^f+(t^m-t^f)\big{)}(1-2r)/2r
+O(\epsilon^2)
\end{split}
\label{eq:freq_diffs}
\end{equation}

\noindent
where $V_{A}=\bar{p}(1-\bar{p})$ is the variance in the frequency of $A$ and $C^d=(\bar{p}+(1-2\bar{p})h^d) s^d +2\alpha_{\Delta}^d$. 
The frequency of Y among male gametes is proportional to the difference in $A$ allele frequency on X- and Y-bearing male gametes, $q=1/2+\alpha_{\Delta}^m(\hat{p}^m_Y-\hat{p}^m_X)+O(\epsilon^3)$.
Without haploid competition or drive ($\alpha_{\Delta}^d=t^d=0$), these results reduce to those of \citet{vanDoorn:2007eu}.

@@

\subsection*{Sex chromosome turnover}
The evolution of a new sex determination system requires that a rare mutant, $m$, at the novel sex-determining locus  increases in frequency when rare. 
The spread of a rare mutant $m$ at the \textbf{M} locus is determined by the leading eigenvalue, $\lambda$, of the system described by the next generation frequency of eggs and sperm carrying the mutation, (\ref{eq:recursions}c), (\ref{eq:recursions}d), (\ref{eq:recursions}g), (\ref{eq:recursions}h), which is an eight equation system. %, evaluated at the above equilibrium.
%Thus, the system describing the change in frequency of the $m$ allele consists of eight recursion equations. 
Dominant neo-Y chromosomes (when $k=0$) or neo-W chromosomes (when $k=1$) are only found in male diploids (neo-Y) or female diploids (neo-W) such that their growth rate ultimately depends only on the change in frequency of $m$-bearing gametes produced by males (for a neo-Y) or by females (for a neo-W). 
Furthermore, if the $m$ allele is fully dominant over the ancestral sex-determining system, phenotypes are not affected by the genotype at the ancestral sex-determining region (\textbf{X} locus). 
Therefore, the invasion of rare mutant neo-Y or neo-W chromosomes can be described by a simpler, two equation system, the largest eigenvalue of which solves the characteristic polynomial

@@

\noindent
where $b$ is the average of the growth rates of the two haplotypes that carry the $m$ allele ($mA$ and $ma$), $b=(\lambda_{mA}^{m}+\lambda_{ma}^{m})/2$, and $c$ also involves the fitness of $m$ alleles when they recombine onto the other \textbf{A} background in a heterozygote, $c=\lambda_{mA}\lambda_{ma}+\rho_{mA}\rho_{ma}$, see table \ref{tab:haplotype_growth}.

\begin{threeparttable}[ht]
\centering
\smallskip
\caption{Parameters determining invasion (equation \ref{eq:charpoly_neoY}) for neo-Y or neo-W chromosomes}
\begin{tabular}{l}
\hline\hline
   \noalign{\vskip 0.5ex}
   neo-Y ($k=0$) \\ [0.5ex] \hline \noalign{\vskip 1ex}
  $\lambda_{mA}=\{p_{X}^{f} w_{A}^{f} w_{A}^{m} w_{AA}^{m}+(1-p_{X}^{f}) w_{a}^{f} w_{A}^{m} w_{Aa}^{m}\alpha^m(1-R)\}/ \bar{w}_{H}^{f} \bar{w}_{H}^{m} \bar{w}^m$ \\ [0.5ex] \noalign{\vskip 0.5ex}
  $\lambda_{ma}=\{(1-p_{X}^{f}) w_{a}^{f} w_{a}^{m} w_{aa}^{m}+p_{X}^{f} w_{A}^{f} w_{a}^{m} w_{Aa}^{m}(1-\alpha^m)(1-R)\}/ \bar{w}_{H}^{f} \bar{w}_{H}^{m} \bar{w}^m $ \\ [0.5ex] \noalign{\vskip 0.5ex}
  $\rho_{mA}=R \{(1-p_{X}^{f}) w_{a}^{f} w_{A}^{m} w_{Aa}^{m} (1-\alpha_{m})\}/ \bar{w}_{H}^{f} \bar{w}_{H}^{m} \bar{w}^m $\\ [0.5ex] \noalign{\vskip 0.5ex}
  $\rho_{ma}=R\{p_{X}^{f} w_{A}^{f} w_{a}^{m} w_{Aa}^{m} \alpha_{m}\}/ \bar{w}_{H}^{f} \bar{w}_{H}^{m} \bar{w}^m$\\ [1ex] \hline 
  \noalign{\vskip 0.5ex}
  neo-W ($k=1$) \** \\ [0.5ex] \hline \noalign{\vskip 1ex}
  $\lambda_{mA}=\{\bar{p}^{m} w_{A}^{m} w_{A}^{f} w_{AA}^{f}+(1-\bar{p}^{m}) w_{a}^{m} w_{A}^{f} w_{Aa}^{f}\alpha^f(1-R)\}/ \bar{w}_{H}^{f} \bar{w}_{H}^{m} \bar{w}^f$ \\ [0.5ex] \noalign{\vskip 0.5ex}
  $\lambda_{ma}=\{(1-\bar{p}^{m}) w_{a}^{m} w_{a}^{f} w_{aa}^{f}+\bar{p}^{m} w_{A}^{m} w_{a}^{f} w_{Aa}^{f}(1-\alpha^f)(1-R)\}/ \bar{w}_{H}^{f} \bar{w}_{H}^{m} \bar{w}^f $ \\ [0.5ex] \noalign{\vskip 0.5ex}
  $\rho_{mA}=R \{(1-\bar{p}^{m}) w_{a}^{m} w_{A}^{f} w_{Aa}^{f} (1-\alpha_{f})\}/ \bar{w}_{H}^{f} \bar{w}_{H}^{m} \bar{w}^f $\\ [0.5ex] \noalign{\vskip 0.5ex}
  $\rho_{ma}=R\{\bar{p}^{m} w_{A}^{m} w_{a}^{f} w_{Aa}^{f} \alpha_{f}\}/ \bar{w}_{H}^{f} \bar{w}_{H}^{m} \bar{w}^f$\\ [1ex] 
  \hline \hline 
   \end{tabular}
      \begin{tablenotes}
      \scriptsize
      \item \** $\bar{p}^{m}=p_{Y}^{m}q+p_{X}^{m}(1-q)$ is the average frequency of the $A$ allele among X- and Y-bearing male gametes (pollen/sperm). 
    \end{tablenotes}
  \label{tab:haplotype_growth}
\end{threeparttable}

Equation \eqref{eq:charpoly_neoY} and table \ref{tab:haplotype_growth} illustrate a number of key points about the invasion of neo-Y and neo-W mutations. 
For a neo-Y, invasion depends on the relative \textcolor{red}{(is this right, is relative fitness divided by mean fitness or difference from 1???)} fitness of $A$-bearing and $a$-bearing male gametes (i.e., in sperm only, not eggs). 
The fitness of male gametes partly depends on the allele carried by the female gamete that they mate with (e.g., $A$ with probability $p_{X}^{f} w_{A}^{f}/\bar{w}_{H}^{f}$).
Similarly, invasion of a neo-W depends on the relative fitness of $A$-bearing and $a$-bearing female gametes. 
However, in the case of a neo-W, the allele carried by the male gamete that they mate with can come from either an X-bearing or a Y-bearing sperm (e.g., $A$ with probability $\bar{p}^{m} w_{A}^{m}/\bar{w}_{H}^{m}$, where $\bar{p}^{m}=p_{Y}^{m}q+p_{X}^{m}(1-q)$). In either case, the zygote will then develop as a female due to the presence of a neo-W. 
By contrast, females that do not carry the neo-W, only result from matings with X-bearing sperm (e.g., matings with $A$-bearing sperm occur with probability $\bar{p}_{X}^{m} w_{A}^{m}/\bar{w}_{H}^{m}$). 
If the \textbf{A} locus is initially linked to the ancestral sex-determining locus, \textbf{X}, (i.e., $r<1/2$) the frequency of the $A$ allele among X- and Y-bearing sperm can differ (equation \ref{eq:freq_diffs}). 
Thus, eggs with and without a neo-W differ in the frequency of alleles they obtain from mating with male gametes.

Previous research suggests, when the ancestral sex-determining locus is linked to a locus that experiences haploid selection (e.g., meiotic drive), a new, unlinked sex-determining locus can invade in order to restore equal sex ratios \citep{Kozielska:2010vm}. 
Our model provides a good opportunity to determine whether Fisherian sex ratio selection provides a useful explanation for the evolution of new sex-determining loci in other contexts. 
Consider, for example, the case where the \textbf{A} locus is linked to the ancestral-SDR ($r<1/2$) and experiences meiotic drive in males only (e.g., during spermatogenesis but not during oogenesis, $\alpha^m \neq 1/2$, $\alpha^f=1/2$). 
We will also disregard haploid competition ($t^f=t^m=0$) such that zygotic sex ratios can only be biased by meiotic drive in males. 
In this case, the zygotic sex ratio can be initially biased only if the ancestral sex-determining system is XY. 
If the ancestral sex-determining system is ZW, the zygotic sex ratio will be 1:1 because diploid sex is determined by the proportion of Z-bearing versus W-bearing eggs (and meiosis in females is fair).
Thus, if the zygotic sex ratio is crucial to the evolution of new genetic sex-determining systems, invasion into ZW and XY systems will be distinct. 
However, we find that invasion of new homogametic systems (XY to XY, or ZW to ZW) and heterogametic systems (XY to ZW or ZY to XY) occur under the same conditions. 
That is, $\lambda_{XY,XY}=\lambda_{ZW,ZW}$ and $\lambda_{XY,ZW}=\lambda_{ZW,XY}$, where $\lambda_{i,j}$ indicates the growth rate of a new mutant giving sex-determining system $j$ in a population in which sex is ancestrally determined by system $i$.

We are particularly concerned with whether or not a rare neo-sex-determining region increases in frequency, which occurs when the largest eigenvalue, $\lambda$, that solves \eqref{eq:charpoly_neoY} is greater than one. 
If the average change in frequency of the two haplotypes that carry the $m$ allele ($Am$ and $am$) is positive, invasion will always occur (if $\left\{(\lambda_{mA}-1)+ (\lambda_{ma}-1) \right\}/2>0$, $\lambda>1$). 
If neither haplotype increases in frequency ($\lambda_{mA}, \lambda_{ma} < 1$ \textcolor{red}{is this notation valid?}), the $m$ allele will not invade. 
Otherwise, the new sex-determining allele increases in frequency on one \textbf{A} background and declines on the other, and invasion requires
\begin{equation}\label{eq:neoYR}
R \left[ 
\frac{p^f_X w^m_a (1-\alpha^m)}{(\lambda_{mA}-1)\bar{w}_{H}^{m}} + 
\frac{(1 - p^f_X) w^m_A \alpha^m \bar{w}_{H}^{m}}{(\lambda_{ma}-1)\bar{w}_{H}^{m}} \right] 
\frac{w^m_{Aa}}{\bar{w}^m} < 1,
\end{equation}

\noindent
for the neo-$Y$, and

\noindent
\begin{equation}\label{eq:neoWR}
R \left[  
\frac{(1 - \bar{p}^m) w^m_{a}}{(\lambda_{mA}-1)\bar{w}_{H}^{m}} + 
\frac{\bar{p}^m w^m_{A}}{(\lambda_{ma}-1)\bar{w}_{H}^{m}} \right] 
\frac{w^f_{Aa}}{\bar{w}^f} < 1
\end{equation}

\noindent
for the neo-$W$.

@@

%<<<<<<< HEAD
%\subsubsection*{Neo-$Y$}
%
%The growth rate of a rare, dominant neo-$Y$ region ($k=0$) is
%\begin{equation}
%\lambda_{Y,XY} = 1 + V_A \frac{(r - R)}{r R} \frac{s_f^2}{\left( s_f + s_m \right)^2} t_m^2,
%\end{equation}
%where $V_A = p_A(1-p_A)$ is the variance at the \textbf{A} locus, $s_f$ and $s_m$ are the respective selection coefficients for $A$ in diploid females and males, and $t_m$ is the selection coefficient for $A$ in gametes/gametophytes from males.
%The neo-Y can therefore invade whenever it is in tighter linkage with the selected locus than is the ancestral sex-determining locus, $r>R$, provided locus \textbf{A} is polymorphic ($V_A>0$) and there is selection among both females ($s_f\neq0$) and gametes/gametophytes from males ($t_m\neq0$).
%=======
%>>>>>>> c89cd9e1d25a76e5bffdd1ee7574884f24c08070
%write in general form without weak selection
%A rare, dominant neo-$Y$ ($k=0$) or neo-$W$ ($k=1$) is always expected to invade an ancestral $XY$ system when the average growth rate of the mutant haplotypes ($Am$ and $am$) is positive, $(g_A + g_a)/2 > 0$. %($g_i = where $\lambda_i$ are the two eigenvalues when $R=r=0$; the \textbf{X} locus does not affect the dynamics).
%When the growth rates of the mutant haplotypes without recombination ($R=0$) are negative, $g_i^* < 0 \; \forall \; i \in \{A,a\}$, where $g_i < g_i^*$, the new sex-determining allele does not invade.

%Otherwise, the new sex-determining allele increases in frequency on one \textbf{A} background and declines on the other, and invasion requires
%\begin{equation}\label{eq:neoYR}
%R \left[ \frac{p^f_X w^m_a (1-\alpha^m)}{g_A^*} + \frac{(1 - p^f_X) w^m_A \alpha^m}{g_a^*} \right] w^m_{Aa} < \nu^m,
%\end{equation}
%for the neo-$Y$, and 
%\begin{equation}\label{eq:neoWR}
%R \left[  \frac{(1 - \bar{p}^m) w^m_{a}}{g_A^*} + \frac{\bar{p}^m w^m_{A}}{g_a^*} \right] w^f_{Aa} < \nu^f
%\end{equation}
%for the neo-$W$.
%Here $\bar{p}^m = (1-q) p^m_X+ q p^m_Y$ is the average frequency of $A$ in gametes produced by males, $w^i_{j}$ is the relative viability fitness of sex $i\in\{m,f\}$ depending on their haploid or diploid genotype at the \textbf{A} locus (with $j\in\{A,a,AA,Aa,aa\}$) and
%$\nu^i = 
%  p^f_X p^m_{\textbf{X}^i} w^m_{a} w^i_{AA} +
%  p^f_X (1-p^m_{\textbf{X}^i}) w^m_{a} w^i_{Aa} +
%  (1-p^f_X) p^m_{\textbf{X}^i} w^m_{a} w^i_{Aa} +
%  (1-p^f_X) (1-p^m_{\textbf{X}^i}) w^m_a w^i_{aa}$,
%with $\textbf{X}^m = Y$ and $\textbf{X}^f = X$, is the mean relative fitness of resident individuals of sex $i$.
%and $g_j^*$ is the growth rate of $X^m$ and $Ym$ haplotypes when on an $A$ ($j=A$)  or $a$ background ($j=a$).
%Although male meiotic drive does not explicitly appear in Equation \ref{eq:neoWR}, it does affect the average frequency of $A$ in gametes from males, $\bar{p}^m$, and thus can play a role in neo-W invasion.
